---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(dplyr)
library(ggplot2)
setwd(
"/Users/abhishekdubey/abhishek-git-repos/datasciencecoursera/reproducible-research/RepData_PeerAssessment1"
)
activity <- read.csv(file = "activity.csv", header = TRUE, stringsAsFactors = FALSE, na.strings = "NA")
activity$date <- as.POSIXct(strptime(with(activity, date), "%Y-%m-%d"))
activityCompleteCases <- activity[complete.cases(activity),]
```

## What is mean total number of steps taken per day?
```{r}
activitySummary <- select(activityCompleteCases, date, steps) %>% group_by(date) %>% summarize(totalSteps = sum(steps), mean = mean(steps), median = median(steps))
hist(activitySummary$totalSteps, col = "red", xlab = "Total Steps", main = "Total Number of Steps Taken Per Day")
```

**Mean and Median of total number of steps taken per day**
```{r}
print(select(activitySummary, date, mean, median), n = nrow(activitySummary))
```


## What is the average daily activity pattern?
```{r}
plot(activityCompleteCases$interval, activityCompleteCases$steps, type = "l", lwd = 2, xlab = "5-minute interval", ylab = "Average number of steps taken", main = "Time Series Plot")
activity.maxSteps <- filter(activityCompleteCases, steps == max(steps))
```

```{r simulation, echo = FALSE}
print("The 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps is as following: ")
print(activity.maxSteps)
```

## Imputing missing values
```{r, echo = FALSE}
number.of.rows.with.missing.values <- sum(!complete.cases(activity))

print(paste0("Total number of missing values in the dataset = ", number.of.rows.with.missing.values))
```

#####We will be using "mean for that 5-minute interval" for imputing(filling-in) missing steps data. In below code, "activity.imputed.data" is the new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activitySummaryByInterval <- select(activityCompleteCases, interval, steps) %>% group_by(interval) %>% summarize(mean = mean(steps))

activity.missing.data.rows <- activity[which(is.na(activity$steps)), ]
activity.not.missing.data.rows <- activity[which(!is.na(activity$steps)), ]
activity.missing.data.rows$steps <- activitySummaryByInterval$mean[match(activitySummaryByInterval$interval, activity.missing.data.rows$interval, nomatch = 0)]

activity.imputed.data <- arrange(rbind(activity.missing.data.rows,activity.not.missing.data.rows), date, interval)
```


## Are there differences in activity patterns between weekdays and weekends?

